A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-22 inst/manuscript/manuscript.tex
more minor updates

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualtions and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between the
    analysis approaches is small when spatially balanced samples are used
    but large when non-spatially balanced samples are used. This suggests
    that the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design and model-based principles (see e.g., Sterba (2009) and
Cicchitelli and Montanari (2012), and see Chan-Golston et al. (2020) for
a Bayesian approach).

Certainly comparisons between design-based and model-based approaches to
spatial data have been studied. But no numerical comparison has been
made between design-based approaches that incorporate spatial
information and model-based approaches. In this manuscript, we compare
design-based approaches that incorporate spatial information to
model-based approaches for finite population spatial data. A finite
population contains a finite number of population units (we assume the
finite number is known); an example is lakes (treated as a whole with
the lake centroid representing location) in the contiguous United
States. Though we manuscript focuses on finite populations, these
comparisons generalize to infinite populations as well. An infinite
population contains an infinite number of population units; an example
is locations within a single lake. In this manuscript we assume the
number of finite population units is known

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples (samples that incorporate spatial
locations of the population units). Spatially balanced samples are
useful because parameter estimates from these samples tend to vary less
than parameter estimates from samples that are not spatially balanced
(Barabesi and Franceschi, 2011; Benedetti et al., 2017; Grafström and
Lundström, 2013; Robertson et al., 2013; Stevens and Olsen, 2004; Wang
et al., 2013). The first spatially balanced sampling algorithm to see
widespread use was the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004). To quantify the spatial balance of
a sample, Stevens and Olsen (2004) proposed loss metrics based on
Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was
developed, several other spatially balanced sampling algorithms emerged,
such as the Local Pivotal Method (Grafström et al., 2012; Grafström and
Matei, 2018), Spatially Correlated Poisson Sampling (Grafström, 2012),
Balanced Acceptance Sampling (Robertson et al., 2013),
Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning Sampling (Robertson et al., 2018). In this
manuscript, we select spatially balanced samples using the Generalized
Random Tessellation Stratified (GRTS) algorithm because it has several
attractive properties. More specifically, the GRTS algorithm
accommodates finite and infinite sampling frames, equal, unequal, and
proportional (to size) inclusion probabilities, legacy (historical)
sampling (Foster et al., 2017), a minimum distance between units in a
sample, and replacement units (replacement units are population units
that can be sampled when a population unit originally selected can no
longer be sampled). The GRTS algorithm selects samples by utilizing a
particular mapping between two-dimensional and one-dimensional space
that preserves proximity relationships. Via this mapping, units in
two-dimensional space are partitioned using a hierarchical address. This
hierarchical address is used to map population units to a
one-dimensional line. On the one dimensional line, each population
unit's line length equals its inclusion probability. Then, a systematic
sample of population units is selected on the line and mapped back to
two-dimensional space, yielding the desired sample. Stevens and Olsen
(2004) provide more technical details.

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th population unit in
the sample and \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample. An estimate of the population mean is
obtained by dividing \(\hat{\tau}_{ht}\) by \(N\), the number of
population units.

It is also important to quantify uncertainty \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically.
Second, these estimators ignore the spatial locations of the population
units. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) and yield narrower confidence intervals compared to
variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next (for more technical details, see Ver Hoef (2008)). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal one). Denoting quantities that are part of the sampled population
units with a subscript \emph{s} and quantities that are part of the
unsampled population units with subscript \emph{u}, let

@@

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the distance between population units), isotropic (independent of
direction), and decay with distance between population units (Cressie,
1993). Henceforth, it is implied that we have made these same
assumptions regarding \(\bm{\delta}\), though Chiles and Delfiner
(1999), pp.~80-93 discuss covariance functions that are not second-order
stationary, not isotropic, or not either. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (for a
thorough list of spatial covariance functions, see Cressie (1993). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013) and random forests (Breiman, 2001), among
others, could also be used to obtain predictions for a mean or total
from spatially correlated responses of a finite population. Compared to
the k-nearest-neighbors and random forest approach, we prefer FBPK
because it is model-based and relies on theoretically-based variance
estimators leveraging the model's spatial covariance structure, whereas
k-nearest-neighbors and random forests use ad-hoc variance estimators
(Ver Hoef and Temesgen, 2013). Additionally, Ver Hoef and Temesgen
(2013) studied compared FBPK, k-nearest-neighbors, and random forests in
a variety of spatial data contexts, and FBPK tended to perform best.

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations: IRS sampling with a design-based
analysis, called ``IRS-Design''; IRS sampling with a model-based
analysis, called ``IRS-Model''; GRTS sampling with a design-based
analysis, called ``GRTS-Design''; GRTS sampling with a model-based
analysis, called ``GRTS-Model''. These combinations are also provided in
Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts, two response
types, and three proportions of dependent random error. The three sample
sizes (\(n\)) were \(n = 50, n = 100,\) and \(n = 200\). Samples were
always selected from a population size (\(N\)) of \(N = 900\). The two
location layouts (of the population units) were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance is 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct
confidence (design-based) or prediction (model-based) intervals. Then we
recorded the bias, squared error, and interval coverage for all
sampling-analysis combinations. After all 2000 trials, we summarized the
long-run performance of the combinations by calculating average bias,
rMS(P)E (root-mean-squared error for the design-based approaches and
root-mean-squared-prediction error for the model-based approaches), and
the proportion of times the true mean is contained in its 95\% interval.
The GRTS algorithm and the local neighborhood variance estimator are
available in the \textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle
et al., 2021). FPBK is available in the \texttt{sptotal}
\textbf{\textsf{R}} package (Higham et al., 2021) and covariance
parameters were estimated using Restricted Maximum Likelihood (Harville,
1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. We will use data from the 2012 National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). Specifically, we will analyze mercury concentration in
lakes. Although we can calculate the true mean mercury concentration
values for the 986 lakes from the 2012 NLA, we will explore whether or
not we obtain an adequately precise estimate for the realized mean
mercury concentration if we sample only 100 of the 986 lakes. For each
of the four familiar sampling-analysis combinations (IRS-Design,
IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or
predict (model-based) the mean mercury concentration and construct 95\%
confidence (design-based) or prediction (model-based) intervals from
this sample of 100 lakes, which we compare to the actual mean from all
986 lakes.

@@

The average bias was nearly zero for all four combinations in all 36
scenarios, so we omit a more detailed summary of those results here.
Tables for average bias in all 36 simulation scenarios are provided in
the supporting information.

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four approaches
from Table \ref{tab:designanalysis} using the random location layout
with ``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
correlation. When there is spatial correlation (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial correlation
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial correlation, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line in each plot represents 95\% coverage.}
  \label{fig:figconf}
\end{figure}

@@

Fig. \ref{fig:merc} shows a map and histogram of mercury concentration.
The map shows mercury concentration exhibits some spatial patterning,
with high mercury concentrations in lakes in the northeast and north
central United States. The histogram shows that mercury concentration is
right-skewed, with most lakes having a low value of mercury
concentration but a few having a much higher concentration. Fig.
\ref{fig:merc} also shows mercury concentration's empirical
semivariogram. The empirical semivariogram can be used as a tool to
visualize spatial dependence. It quantifies the halved squared
differences (semivariance) among mercury concentration at different
distances apart. When a process is spatially correlated (has spatial
dependence), the semivariance tends to be smaller at small distances and
larger at large distances. The empirical semivariogram in Fig.
\ref{fig:merc} suggests that mercury concentration is exhibits spatial
dependence. Lastly we note that the realized mean mercury concentration
in the 986 NLA lakes is 103.2 ng / g.

@@

@@

@@

In the supporting information, we provide tables presenting summary
statistics for all 36 simulation scenarios.

\hypertarget{author-contributions}{%
\section*{Author Contributions}\label{author-contributions}}
\addcontentsline{toc}{section}{Author Contributions}

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualations and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between
    these approaches is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used. This suggests that
    the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the sampling or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design and model-based principles (see e.g., Sterba (2009) and
Cicchitelli and Montanari (2012), and see Chan-Golston et al. (2020) for
a Bayesian approach).

Certainly comparisons between design-based and model-based approaches to
spatial data have been studied. But no numerical comparison has been
made between design-based approaches that incorporate spatial
information and model-based approaches. In this manuscript, we compare
design-based approaches that incorporate spatial information to
model-based approaches for finite population spatial data. A finite
population contains a finite number of population units (we assume the
finite number is known); an example is lakes (treated as a whole with
the lake centroid representing location) in the contiguous United
States. Though we focus on finite populations, these comparisons
generalize to infinite populations as well. An infinite population
contains an infinite number of population units; an example is locations
within a single lake.

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples (samples that incorporate spatial
locations of the population units). Spatially balanced samples are
useful because parameter estimates from these samples tend to vary less
than parameter estimates from samples that are not spatially balanced
(Barabesi and Franceschi, 2011; Benedetti et al., 2017; Grafström and
Lundström, 2013; Robertson et al., 2013; Stevens and Olsen, 2004; Wang
et al., 2013). The first spatially balanced sampling algorithm to see
widespread use was the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004). To quantify the spatial balance of
a sample, Stevens and Olsen (2004) proposed loss metrics based on
Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was
developed, several other spatially balanced sampling algorithms emerged,
including the Local Pivotal Method (Grafström et al., 2012; Grafström
and Matei, 2018), Spatially Correlated Poisson Sampling (Grafström,
2012), Balanced Acceptance Sampling (Robertson et al., 2013),
Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning Sampling (Robertson et al., 2018). In this
manuscript, we select spatially balanced samples using the Generalized
Random Tessellation Stratified (GRTS) algorithm because it has several
attractive properties: the GRTS algorithm accommodates finite and
infinite sampling frames, equal, unequal, and proportional (to size)
inclusion probabilities, legacy (historical) sampling (Foster et al.,
2017), a minimum distance between units in a sample, and replacement
units (replacement units are population units that can be sampled when a
population unit originally selected can no longer be sampled). The GRTS
algorithm selects samples by utilizing a particular mapping between
two-dimensional and one-dimensional space that preserves proximity
relationships. Via this mapping, units in two-dimensional space are
partitioned using a hierarchical address. This hierarchical address is
used to map population units to a one-dimensional line. On the one
dimensional line, each population unit's line length equals its
inclusion probability. Then, a systematic sample of population units is
selected on the line and mapped back to two-dimensional space, yielding
the desired sample. Stevens and Olsen (2004) provide more technical
details.

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th population unit in
the sample, \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample, and \(n\) is the sample size. An estimate
of the population mean is obtained by dividing \(\hat{\tau}_{ht}\) by
\(N\), the number of population units.

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically.
Second, these estimators ignore the spatial locations of the population
units. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) and yield narrower confidence intervals compared to
variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next -- for technical details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal one). Denoting quantities that are part of the sampled population
units with a subscript \emph{s} and quantities that are part of the
unsampled population units with a subscript \emph{u}, let

@@

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the distance between population units), isotropic (independent of
direction), and decay with distance between population units (Cressie,
1993). Henceforth, it is implied that we have made these same
assumptions regarding \(\bm{\delta}\), though Chiles and Delfiner
(1999), pp.~80-93 discuss covariance functions that are not second-order
stationary, not isotropic, or not either. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (Cressie
(1993) provides a thorough list of spatial covariance functions). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013) and random forest (Breiman, 2001), among
others, could also be used to obtain predictions for a mean or total
from finite population spatial data. Compared to the k-nearest-neighbors
and random forest approach, we prefer FBPK because it is model-based and
relies on theoretically-based variance estimators leveraging the model's
spatial covariance structure, whereas k-nearest-neighbors and random
forests use ad-hoc variance estimators (Ver Hoef and Temesgen, 2013).
Additionally, Ver Hoef and Temesgen (2013) studied compared FBPK,
k-nearest-neighbors, and random forest in a variety of spatial data
contexts, and FBPK tended to perform best.

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations: Independent Random Sampling (IRS) with a
design-based analysis, called ``IRS-Design'' (IRS samples are not
spatially balanced because population units are selected independent of
their spatial locations); IRS with a model-based analysis, called
``IRS-Model''; GRTS sampling with a design-based analysis, called
``GRTS-Design''; and GRTS sampling with a model-based analysis, called
``GRTS-Model''. These combinations are also provided in Table
\ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance is 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Then we recorded the bias, squared error, and interval coverage for all
sampling-analysis combinations. After all 2000 trials, we summarized the
long-run performance of the combinations by calculating average bias,
rMS(P)E (root-mean-squared error for the design-based approaches and
root-mean-squared-prediction error for the model-based approaches), and
the proportion of times the true mean is contained in its 95\% interval.
The GRTS algorithm and the local neighborhood variance estimator are
available in the \texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle
et al., 2021). FPBK is available in the \texttt{sptotal}
\textbf{\textsf{R}} package (Higham et al., 2021) and covariance
parameters were estimated using Restricted Maximum Likelihood (Harville,
1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. We will use data from the 2012 National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). Specifically, we will analyze mercury concentration in
lakes. Although we can calculate the true mean mercury concentration
values for the 986 lakes from the 2012 NLA, we will explore whether or
not we obtain an adequately precise estimate for the realized mean
mercury concentration if we sample only 100 of the 986 lakes. For each
of the four familiar sampling-analysis combinations (IRS-Design,
IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or
predict (model-based) the mean mercury concentration and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals
from this sample of 100 lakes, which we compare to the true mean mercury
concentration from all 986 lakes.

@@

The average bias was nearly zero for all four sampling-analysis
combinations in all 36 scenarios, so we omit a more detailed summary of
those results here. Tables for average bias in all 36 simulation
scenarios are provided in the supporting information.

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
covariance. When there is spatial covariance (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial covariance
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial covariance, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line represents 95\% coverage.}
  \label{fig:figconf}
\end{figure}

@@

Fig. \ref{fig:merc} shows a map and histogram of mercury concentration
in all 986 NLA lakes. The map shows mercury concentration exhibits some
spatial patterning, with high mercury concentrations in the northeast
and north central United States. The histogram shows that mercury
concentration is right-skewed, with most lakes having a low value of
mercury concentration but a few having a much higher concentration. Fig.
\ref{fig:merc} also shows mercury concentration's empirical
semivariogram. The empirical semivariogram can be used as a tool to
visualize spatial dependence. It quantifies the halved squared
differences (semivariance) among mercury concentration at different
distances apart. When a process has spatial covariance (exhibits spatial
dependence), the semivariance tends to be smaller at small distances and
larger at large distances. The empirical semivariogram in Fig.
\ref{fig:merc} suggests that mercury concentration is exhibits spatial
dependence. Lastly we note that the realized mean mercury concentration
in the 986 NLA lakes is 103.2 ng / g.

@@

\hypertarget{author-contribution-statement}{%
\section*{Author Contribution
Statement}\label{author-contribution-statement}}
\addcontentsline{toc}{section}{Author Contribution Statement}

@@

If the manuscript is accepted, this repository will be archived in
Zenodo.

@@

In the supporting information, we provide tables of summary statistics
for all 36 simulation scenarios.
